---
title: 'Reproducible Research: Course Project 1'
author: "Bibek Karki"
date: "01/15/2017"
output: html_document
---

```{r globalOptions, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This project makes use of [Activity Monitoring Data] (https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) dataset which was collected using a personal activity monitoring device. The data includes number of steps taken at every 5 minute interval over a two month period (October - November, 2012). The objective of the project is to explore the the walking behavior of related person using the dataset.


## Loading and preprocessing the data

The current directory is expected to be in **RepData_PeerAssessment1**, the [repository forked] (https://github.com/rdpeng/RepData_PeerAssessment1) for the project.

```{r loadData}
if(!file.exists("activity.csv")) {
        unzip("activity.zip")
}
AMD <- read.csv("activity.csv")
```

## What is mean total number of steps taken per day?



## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
