---
title: 'Reproducible Research: Course Project 1'
author: "Bibek Karki"
date: "01/15/2017"
output: html_document
---

```{r globalOptions, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This project makes use of [Activity Monitoring Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) dataset which was collected using a personal activity monitoring device. The data includes number of steps taken at every 5 minute interval over a two month period (October - November, 2012). The objective of the project is to explore the the walking behavior of the person using the dataset.



## Loading and preprocessing the data

The current directory is expected to be in **RepData_PeerAssessment1**, which is the [repository forked](https://github.com/rdpeng/RepData_PeerAssessment1) for the project.

```{r loadData}
# Loading the data
if(!file.exists("activity.csv")) {
        unzip("activity.zip")
}

if (!exists("AMD")) {
        AMD <- read.csv("activity.csv")
}

# Preprocessing
library(ggplot2)
```



## What is mean total number of steps taken per day?

**Step 1:** Calculate the total steps taken per day.

**Step 2:** Plot the histogram distribution of total number of steps per day.

```{r totalStepsPerDay}
## Step 1
totalSteps <- aggregate(steps ~ date, data = AMD, FUN = sum)

## Step 2
ggplot(data = totalSteps, mapping = aes(steps)) +
        geom_histogram(bins = 9, col = "steelblue", fill = rgb(0.00,0.33,0.67)) + 
        labs(title = "Distribution of Total Steps per Day", x = "Steps/day", y = "Frequency") +
        theme_bw(base_family = "Times")
```

**Step 3:** Calculate the mean and median steps per day

```{r medianMean}
as.integer(mean(totalSteps$steps))
as.integer(median(totalSteps$steps))
```



## What is the average daily activity pattern?

**Step 1:** Calculate average number of steps for each time interval

**Step 2:** Show the result in line plot

```{r averageStepsPerInterval}
## Step 1
averageSteps <- aggregate(steps~interval,data = AMD, FUN = mean)

## Step 2
ggplot(data = averageSteps, mapping = aes(x = interval, y = steps)) +
        geom_line(lty = 1, col = rgb(0.00,0.33,0.67)) + 
        labs(title = "Average Steps per Time Interval", x = "Time Interval", y = "Average Number of Steps") +
        theme_bw(base_family = "Times")
```

**Step 3:** Which 5-minute interval has the largest average steps
```{r maxAvgSteps}
## Step 3
averageSteps$interval[which.max(averageSteps$steps)]
```



## Imputing missing values

**Step 1:** Calculate the total number of missing values

```{r missingValues}
## Step 1
sum(is.na(AMD$steps))
```



## Are there differences in activity patterns between weekdays and weekends?
